Computational biology in the 21st century: Algorithms that scale

### 2015-07-29 main.tex
Figure and table cleanup

\usepackage{caption}
\usepackage{subfig}
\captionsetup[subfigure]{margin=5pt}

@@

% YWY: Take spell-check lines out of figure axes labels! (TeraFLOPS, KiloBase)
\begin{figure}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/moore.png}
\label{fig:moore}
}
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{(a) Moore's and (b) Kryder's laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure}

@@

% YWY: Take spell-check lines out of figure labels! (bp, Samtools)
\begin{figure}[htb!]
\centering
\subfloat[`Shotgun' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[width=2.6in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of `variant-calling' analysis]{
\includegraphics[width=2.6in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=4.0in]{assets/ngs-pipeline.png}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure}

@@

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Any high-dimensional data set, such as amino acid sequences of a bounded 
length, can be described as inhabiting a vector space.
For example, given an alphabet of 20 amino acids and a length bounded at 
500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
in length, proteins are typically described as comprising \emph{domains} that 
evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
could
enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
of July 
2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
This surprisingly lower number is not simply due to undersampling; thanks to 
evolution, only those protein
sequences which exhibit useful biological function should survive, and most
random sequences of amino acids would not be expected to form stable structures.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:clusters}).
It seems that physical laws have constrained the data to a particular subspace
of the enumerable Cartesian space.

\begin{figure}[htb!]
\centering
\subfloat[Cartoon depiction of points in an arbitrary high-dimensional space, 
as might arise from genomes generated by mutation and selection during the 
course of evolution. 
Although high-dimensional locally, at the global scale of covering spheres, the 
data cloud looks nearly 1-dimensional, which enables entropy-scaling of 
similarity search. Clusters cover the data points but do not cover unoccupied 
regions of space.]{
\includegraphics[width=2.8in]{assets/treepoints.pdf}
\label{fig:clusters}
}
\subfloat[A query (green triangle) and two concentric search radii (red 
circles) around it. Thanks to low fractal dimension, the large circle does
not contain vastly more points than the small circle.]{
\includegraphics[width=2.8in]{assets/treepoints-fractal.png}
\label{fig:fractal}
}
\caption[Data spaces.]{A cartoon illustration of a biological data space.}
\label{fig:dataspace}
\end{figure}

One key insight related to redundancy
%as to why the compressive omics approach allows computation 
%that scales sublinearly in the size of the data 
is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number $k$ 
of clusters needed to cover $D$ is bounded by $N_{r_c} (D)$, the metric 
entropy, which is relatively small compared to
$|D|$, the number of entries in the database (Figure~\ref{fig:clusters}).
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

A second key insight is that the biological data sets have low \emph{fractal dimension}~\cite{yu2015entropy}.
That is, within some range of radii $r_1$ and $r_2$ about an arbitrary point
in the database $D$, the fractal dimension $d$ is
$d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$, where $n_1$ and $n_2$ are the
number of points within $r_1$ and $r_2$ respectively (Figure~\ref{fig:fractal}).

Cluster-based search, as exemplified by ``compressive omics,'' can perform
approximate search within a radius $r$ of a query $q$ on a database $D$ with 
fractal dimension $d$ and metric entropy $k$ at the scale $r_c$ in time
proportional to
\[
    O\Bigg(
    \underbrace{k}_{\textrm{metric entropy}} +
    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
     \Bigg) .
\]

@@

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets.
NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\begin{tabular}{ccc}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein structure (PDB) & 10:1 & 2.5\\
\hline
Chemical structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

>>>>

\usepackage{array}
\usepackage{caption}
\usepackage{subfig}
\captionsetup[subfigure]{margin=5pt}

@@

% YWY: Take spell-check lines out of figure axes labels! (TeraFLOPS, KiloBase)
\begin{figure*}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/moore.png}
\label{fig:moore}
}
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{(a) Moore's and (b) Kryder's laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure*}

@@

\begin{figure*}[htb!]
\centering
\subfloat[`Shotgun' sequencing breaks DNA molecules into many short fragments, which are read from one or both ends in the form of {\em reads,\/} and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[height=2.6in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of `variant-calling' analysis]{
\includegraphics[height=2.6in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=4.0in]{assets/ngs-pipeline.png}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure*}

@@

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Any high-dimensional data set, such as amino acid sequences of a bounded 
length, can be described as inhabiting a vector space.
For example, given an alphabet of 20 amino acids and a length bounded at 
500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
in length, proteins are typically described as comprising \emph{domains} that 
evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
could
enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
of July 
2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
This surprisingly lower number is not simply due to undersampling; thanks to 
evolution, only those protein
sequences which exhibit useful biological function should survive, and most
random sequences of amino acids would not be expected to form stable structures.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
It seems that physical laws have constrained the data to a particular subspace
of the enumerable Cartesian space.

\begin{figure}[htb!]
\centering
\includegraphics[width=2.8in]{assets/treepoints-fractal.png}
\caption{Cartoon depiction of points in an arbitrary high-dimensional space, 
as might arise from genomes generated by mutation and selection during the 
course of evolution. 
Although high-dimensional locally, at the global scale of covering spheres, the 
data cloud looks nearly 1-dimensional, which enables entropy-scaling of 
similarity search. Clusters cover the data points but do not cover unoccupied 
regions of space.
The green triangle represents a query, with two concentric search radii (red 
circles) around it. Thanks to low fractal dimension, the large circle does
not contain vastly more points than the small circle.}
\label{fig:dataspace}
\end{figure}

One key insight related to redundancy
%as to why the compressive omics approach allows computation 
%that scales sublinearly in the size of the data 
is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number $k$ 
of clusters needed to cover $D$ is bounded by $N_{r_c} (D)$, the metric 
entropy, which is relatively small compared to
$|D|$, the number of entries in the database (Figure~\ref{fig:dataspace}).
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

A second key insight is that the biological data sets have low \emph{fractal dimension}~\cite{yu2015entropy}.
That is, within some range of radii $r_1$ and $r_2$ about an arbitrary point
in the database $D$, the fractal dimension $d$ is
$d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$, where $n_1$ and $n_2$ are the
number of points within $r_1$ and $r_2$ respectively (Figure~\ref{fig:dataspace}).

Cluster-based search, as exemplified by ``compressive omics,'' can perform
approximate search within a radius $r$ of a query $q$ on a database $D$ with 
fractal dimension $d$ and metric entropy $k$ at the scale $r_c$ in time
proportional to
\begin{displaymath}
    O\Bigg(
    \underbrace{k}_{\textrm{metric entropy}} +
    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
     \Bigg) .
\end{displaymath}

@@

\newcolumntype{H}[1] {%
>{\raggedright}%
p{#1}}

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets.
NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\tabcolsep=0.11cm
\begin{tabular}{H{3.0cm}p{2.2cm}p{1.8cm}}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide~sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein~sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein~structure (PDB) & 10:1 & 2.5\\
\hline
Chemical~structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}
